Stochastic trace estimation for parameter-dependent matrices applied to spectral density approximation

### 2025-02-25 paper/spectraldensity.tex
address review comments, clean up equations, create n_c=2 plot

Choosing the smoothing parameter $\sigma$ results in the following trade off: A larger $\sigma$ usually makes it  easier to approximate the matrix function $g_{\sigma}(t\mtx{I}_n - \mtx{A})$, whereas a smaller $\sigma$ allow one to stay closer to the original spectral density $\phi$. To quantify the latter, one can measure the error between $\phi$ and its smoothened version $\phi_{\sigma}$ by
\begin{equation}
    \sup_{f \in \mathcal{S}} \int_{a}^{b} f(t) (\phi(t) - \phi_{\sigma}(t))~\mathrm{d}t,
    \label{equ:error-metric}
\end{equation}
where $\mathcal{S}$ represents an appropriately chosen space of test functions. Among others, the reference~\cite{lin-2016-approximating-spectral} uses $\mathcal{S} = \{ f: f(t) \equiv g_{\sigma}(t - s), s \in [a, b]\}$, \cite{chen-2021-analysis-stochastic} uses $\mathcal{S} = \{f : f(t) = \Theta(s - t), s \in [a, b] \}$ with the Heaviside step function $\Theta$, and \cite{braverman-2022-sublinear-time,bhattacharjee-2024-improved-spectral} use $\mathcal{S} = \{f : |f(t) - f(s)| \leq |t - s| \}$ for which the error \refequ{equ:error-metric} is equivalent to the Wasserstein-1 distance between $\phi$ and $\phi_{\sigma}$.
When choosing 
the last metric,
assuming that the eigenvalues are somewhat uniformly distributed and $\sigma$ is rather small, a simple calculation allows one to conclude that the smoothing error \refequ{equ:error-metric} is of the order $\sigma$. Consequently, to ensure that the approximation deviates by at most a factor of $\varepsilon > 0$, one needs to choose $\sigma \approx \varepsilon$.

@@

\begin{lemma}{Chebyshev approximation error}{chebyshev-error}
    The Chebyshev approximation $g_{\sigma}^{(m)}$ of the Gaussian smoothing kernel $g_{{\sigma}}$ from  \refequ{equ:smoothing-kernel} satisfies
    \begin{equation}
        \sup_{s,t \in [-1, 1]} \big| g_{\sigma}(t - s) - g_{\sigma}^{(m)}(t - s) \big| \leq 2\sqrt{\frac{2e}{\pi}} \frac{1}{\sigma^2} (1 + \sigma)^{-m} =: E_{\sigma, m}
        \label{equ:chebyshev-interpolation-sup-error-kernel}
        %\lVert  \phi_{\sigma} - \phi_{\sigma}^{(m)} \rVert _1 &\leq \frac{2\sqrt{2}}{\sigma^2} (1 + \sigma)^{-m}.
        %\label{equ:2-chebyshev-interpolation-error}
    \end{equation}
    for every degree $m \in \mathbb{N}$ and smoothing parameter $\sigma > 0$.
\end{lemma}
\begin{proof}
    The proof of this result is standard. % and follows the one given in the arXiv preprint of \cite[Theorem 2]{lin-2017-randomized-estimation}, but attains a slightly different dependence on $\sigma$. 
    Let $\mathcal{E}_{\chi} \subset \mathbb{C}$ denote the Bernstein ellipse  with foci $\{-1, +1\}$ and the sum of semi-axes equal to $\chi > 1$. By\cite[Theorem 8.2]{trefethen-2020-approximation-theory},
    \begin{equation}
        \sup_{s \in [-1, 1]} \big| g_{\sigma}(t - s) - g_{\sigma}^{(m)}(t - s) \big| \leq \frac{4}{\chi^m (\chi - 1)} \max_{z \in \mathcal{E}_{\chi}} |g_{\sigma}(t - z)|,
        \label{equ:bernstein-bound}
    \end{equation}
    where used the analyticity of $g_{\sigma}(t- \cdot)$.
    For $z = x + \mathrm{i} y \in \mathcal{E}_{\chi}$, we have
    \begin{equation}
    | g_{\sigma}(t - z) | 
    = \frac{1}{\sigma \sqrt{2 \pi}} \big| e^{-\sfrac{(t - z)^2}{2\sigma^2}} \big|
    = \frac{1}{\sigma \sqrt{2 \pi}} e^{-\sfrac{(t - x)^2}{2\sigma^2}}e^{\sfrac{y^2}{2\sigma^2}}
    \le \frac{1}{\sigma \sqrt{2 \pi}} e^{\sfrac{(\chi - \chi^{-1})^2}{8 \sigma^2}},
    %\leq \frac{1}{n \sigma \sqrt{2 \pi}} \max_{x + iy \in \mathcal{E}_{\chi}} e^{\sfrac{y^2}{2\sigma^2}} 
    \end{equation}
    where the last inequality uses 
    $e^{-\sfrac{(t - x)^2}{2\sigma^2}} \leq 1$ and $|y| \le (\chi - \chi^{-1}) / 2$.
    Choosing $\chi = 1 + \sigma$ and using $\chi - \chi^{-1} \leq 2\sigma$ completes the proof of~\refequ{equ:chebyshev-interpolation-sup-error-kernel}.
\end{proof}

@@

\begin{lemma}{$L^1$-error of non-negative approximate smooth spectral density}{non-negative-chebyshev-error}
Let $\phi_{\sigma}$ be the smoothed spectral density defined in~\refequ{equ:smooth-spectral-density}
for a symmetric matrix $\mtx{A} \in \mathbb{R}^{n \times n}$ with spectrum contained in $[-1, 1]$. Then the approximation $\underline{\phi}_{\sigma}^{(m)}$ defined in~\refequ{equ:non-negative-expanded-density} satisfies
    \begin{equation}
        \int_{-1}^{1} \big| \phi_{\sigma}(t) - \underline{\phi}_{\sigma}^{(m)}(t) \big|~\mathrm{d}t \leq 4 \sqrt{2} \left(1 + \sigma \sqrt{\pi} \cdot E_{\sqrt{2}\sigma, \sfrac{m}{2}}\right) E_{\sqrt{2}\sigma, \sfrac{m}{2}} =: \underline{E}_{\sigma, m}
        \label{equ:chebyshev-interpolation-sup-error-kernel-nonneg}
        %\lVert  \phi_{\sigma} - \phi_{\sigma}^{(m)} \rVert _1 &\leq \frac{2\sqrt{2}}{\sigma^2} (1 + \sigma)^{-m}.
        %\label{equ:2-chebyshev-interpolation-error}
    \end{equation}
    for every \emph{even} degree $m \in \mathbb{N}$, smoothing parameter $\sigma > 0$, and with $E_{\sigma, m}$ defined in \reflem{lem:chebyshev-error}.
\end{lemma}
\begin{proof}
    Using that the inequality $| a - b | \leq (2 \sqrt{a} + | \sqrt{a} - \sqrt{b} |)  | \sqrt{a} - \sqrt{b} |$ holds for any non-negative numbers $a, b$, it follows that
    \begin{equation}
        \big| g_{\sigma} - \underline{g}_{\sigma}^{(m)} \big| \leq \big( 2 \big| \sqrt{g_{\sigma}} \big| + \big| \sqrt{g_{\sigma}} - \sqrt{g_{\sigma}}^{(\sfrac{m}{2})} \big| \big) \big| \sqrt{g_{\sigma}} - \sqrt{g_{\sigma}}^{(\sfrac{m}{2})} \big|.
    \end{equation}
    Because of $\sqrt{g_{\sigma}} = \sqrt{2 \sigma \sqrt{2 \pi}} \cdot g_{\sqrt{2}\sigma}$,
we can apply \reflem{lem:chebyshev-error} with the substitutions
    $\sigma \gets \sqrt{2}\sigma$ and $m \gets m/2$ to bound 
    $\big| \sqrt{g_{\sigma}} - \sqrt{g_{\sigma}}^{(\sfrac{m}{2})} \big|$.
    Together with  $|\sqrt{g_{\sigma}}| \leq 1/\sqrt{\sigma \sqrt{2 \pi}}$, we get
    \begin{align}
        \big| g_{\sigma}(t - s) - \underline{g}_{\sigma}^{(m)}(t - s) \big| \notag 
        &\leq \Big( 2 \frac{1}{\sqrt{ \sigma \sqrt{2\pi}}} + \sqrt{2 \sigma \sqrt{2 \pi}} \cdot E_{\sqrt{2}\sigma, \sfrac{m}{2}}\Big)\sqrt{2 \sigma \sqrt{2 \pi}} \cdot E_{\sqrt{2}\sigma, \sfrac{m}{2}} \notag \\
        &= 2\sqrt{2} \big(1 + \sigma \sqrt{\pi} \cdot E_{\sqrt{2}\sigma, \sfrac{m}{2}}\big) E_{\sqrt{2}\sigma, \sfrac{m}{2}}
    \end{align}
    for every $s,t \in [-1, 1]$.
    Bounding the integral as in~\refequ{equ:chebyshev-interpolation-spectral-density} then yields the desired result.
\end{proof}

@@

%\textcolor{red}{??? I don't understand this derivation. Don't you think that~\refequ{equ:discrete-cosine-transform} only holds approximately for the exact $g_\sigma$ and it should be something like $g_\sigma \approx g^{(m)} = ...$. I think there is a major confusion of two different approaches here:
%\begin{itemize}
% \item The Chebyshev \emph{approximation} of $f$ is the $\ell^2$-orthogonal projection of $f$ on the first $m+1$ Chebyshev polynomials. The coefficients are computed from $\ell^2$ inner products between the Chebyshev polynomials and $f$. This is probably what Lin is doing. 
% \item The Chebyshev \emph{approximation} used by Chebfun and possibly others proceeds differently. It first replaces (or rather pretends to replace) $f$ by the interpolation of $f$ in (some) Chebyshev nodes. This interpolation admits an exact truncated approximation and the coefficients can be computed from a Fourier transform (or even DCT?) of the interpolation data. This might match what is written below, but I am not sure whether the interpolation nodes are the same. Please check with Trefethen's book.
%\end{itemize}
%There are two issues now
%\begin{enumerate}
% \item I am now unsure how $g_\sigma^{(m)}$ is precisely defined. This should be spelled out clearer in the beginning of Section 3.2.
% \item There could be a mismatch between the $g_\sigma^{(m)}$ of Lemma 3.1 and the one we are actually using. This mismatch can most likely be resolved, because a result of the form of Lemma 3.1 also holds for the Chebyshev interpolation. If it is easy to resolve, please do so. If not, at least point this out here. Have a look at Trefethen's book; it might have the right error bound.
%\end{enumerate}
%???
It is well known that the coefficients of the Chebyshev approximation \refequ{equ:matrix-approximation} can --- in exact arithmetic --- be computed by inverting the (type I) discrete cosine transform (DCT)
\begin{equation}
    g_{\sigma}(t - \cos(s_i)) = \sum_{l=0}^{m} \mu_l(t) \cos\left(\frac{\pi i l}{m} \right)
    \label{equ:discrete-cosine-transform}
\end{equation}
between the approximation coefficients $\mu_l(t),~l=0,\dots,m$ and the function evaluations at the Chebyshev points $s_i = \cos(\pi i / m),~i=0,\dots,m$ \cite{baszenski-1997-fast-polynomial, plonka-2018-numerical-fourier, trefethen-2020-approximation-theory}. This can be accomplished in $\mathcal{O}(m \log(m))$ operations.%}

@@

We start with the first term in the definition~\refequ{equ:nystrompp-trace-estimator} of $\Nystrpp{\mtx{\Omega}}{\mtx{\Psi}}$, which requires evaluating the trace of the 
Nyström approximation \refequ{equ:nystrom-approximation}:
\begin{equation}
    \frac{1}{n} \Trace \big((\mtx{B}(t) \mtx{\Omega}) (\mtx{\Omega}^{\top} \mtx{B}(t) \mtx{\Omega})^{\dagger} (\mtx{B}(t) \mtx{\Omega})^{\top} \big) =
    \frac{1}{n} \Trace \big((\underbrace{\mtx{\Omega}^{\top} \mtx{B}(t) \mtx{\Omega}}_{=: \mtx{K}_1(t)})^{\dagger} (\underbrace{\mtx{\Omega}^{\top} \mtx{B}(t)^2 \mtx{\Omega}}_{=: \mtx{K}_2(t)}) \big).
    %\frac{1}{n} \Trace((\mtx{\Omega}^{\top} \underline{g}_{\sigma}^{(m)}(t \mtx{I}_n - \mtx{A}) \mtx{\Omega})^{\dagger} (\mtx{\Omega}^{\top} \underline{g}_{\sigma}^{(m)}(t \mtx{I}_n - \mtx{A})^2 \mtx{\Omega})),
    \label{equ:cyclic-property}
\end{equation}
where we used the invariance of the trace under cyclic permutations.
Inserting the Chebyshev approximation  $\mtx{B}(t) = \underline{g}_{\sigma}^{(m)}(t \mtx{I}_n - \mtx{A})$ gives
\begin{equation}
    \mtx{K}_1(t) = \mtx{\Omega}^{\top} \underline{g}_{\sigma}^{(m)}(t \mtx{I}_n - \mtx{A}) \mtx{\Omega} = \sum_{l=0}^{m} \underline{\mu}_l(t) (\mtx{\Omega}^{\top} T_l(\mtx{A}) \mtx{\Omega}).
\end{equation}
Observing that $\mtx{\Omega}^{\top} T_l(\mtx{A}) \mtx{\Omega}$ is independent of $t$ allows us to precompute these (small) $n_{\mtx{\Omega}} \times n_{\mtx{\Omega}}$ matrices for $l=0, \dots, m$ using the Chebyshev recurrence \refequ{equ:chebyshev-recurrence}. For each value of $t$, we then only need to sum these matrices (multiplied by the corresponding coefficient $\underline{\mu}_l(t)$).

The squaring of Chebyshev approximations discussed in \refsec{sec:preservnonneg} allows us to express $\mtx{B}(t)^2 = \underline{g}_{\sigma}^{(m)}(t \mtx{I}_n - \mtx{A})^2 = \sum_{l=0}^{2m} \underline{\nu}_l(t) T_l(\mtx{A})$, where the coefficients $\underline{\nu}_l(t)$ can be cheaply computed from $\underline{\mu}_l(t)$ using~\refalg{alg:chebyshev-squaring}. Hence, if we precompute
$\mtx{\Omega}^{\top} T_l(\mtx{A}) \mtx{\Omega}$
also for $l=m+1, \dots, 2m$, this allows us to evaluate $\mtx{K}_2(t)$ cheaply, in an analogous way to $\mtx{K}_1(t)$. Alternatively, \cite{lin-2017-randomized-estimation} proposes to instead approximate the squared smoothing kernel $g_{\sigma}^2$ in a separate "auxiliary" Chebyshev series truncated after term $2m$. While this direct approximation of $g_{\sigma}^2$ might be more accurate than our squaring approach, it is \emph{inconsistent} in the sense that numerically computing the truncated Chebyshev series of $g_{\sigma}^2$ gives --- in general --- not the same result as squaring the truncated Chebyshev series of $g_{\sigma}$. We have observed that meticulously ensuring consistency between the approximations of $g_{\sigma}$ and $g_{\sigma}^2$ makes a visible difference in terms of accuracy (cf. \reffig{fig:interpolation-issue}). This consistency will also allow us to derive bounds for the estimator \refequ{equ:chebyshev-nystrom-formula}.

\begin{figure}[ht]
    \centering
    \input{plots/interpolation.pgf}
    \caption{Accuracy when computing the estimate \refequ{equ:cyclic-property} in three different ways. \emph{Inconsistent}: Applying an auxiliary truncated Chebyshev series to $g_{\sigma}^2$, as proposed in~\cite{lin-2017-randomized-estimation}. \emph{Consistent}: Squaring the Chebyshev approximation $g_{\sigma}^{(m)}$ using \refalg{alg:chebyshev-squaring}. \emph{Non-negative consistent}: Squaring the non-negative Chebyshev approximation $\underline{g}_{\sigma}^{(m)}$ with~\refalg{alg:chebyshev-squaring}. As input matrix we use the $100 \times 100$ matrix $\mtx{A}$ coming from a two-dimensional finite difference discretization of the Hamiltonian defined in~\refequ{equ:5-experiments-electronic-hamiltonian} on a $10 \times 10$ grid. We fix $n_{\mtx{\Omega}} = 80$ to make sure the low-rank approximation is accurate, set $\sigma = 0.005$, and compute the $L^1$ error from the approximation of the spectral density at $n_t = 100$ uniformly spaced values of $t$ in the interval $[-1, 1]$.}
    \label{fig:interpolation-issue}
\end{figure}

@@

\begin{figure}[ht]
    \centering
    \input{plots/zerocheck.pgf}
    \caption{The difference the non-zero check can make when approximating a spectral density using \refalg{alg:nystrom-chebyshev-pp}. As input matrix we use the $1000 \times 1000$ matrix $\mtx{A}$ coming from a three-dimensional finite difference discretization of the Hamiltonian defined in~\refequ{equ:5-experiments-electronic-hamiltonian} on a $10 \times 10$ grid. We ran the Chebyshev-Nyström++ method with and without estimating if the matrix function is zero before taking its pseudo-inverse. We let $m=2000$, $n_{\mtx{\Omega}}=80$, $n_{\mtx{\Psi}}=0$, and $\sigma = 0.004$. We compute the $L^1$ error from the approximation of the spectral density at $n_t = 100$ uniformly spaced values of $t$ in the interval $[-1, 1]$.}
    \label{fig:zerocheck}
\end{figure}

>>>>

Choosing the smoothing parameter $\sigma$ results in the following trade off: A larger $\sigma$ usually makes it  easier to approximate the matrix function $g_{\sigma}(t\mtx{I}_n - \mtx{A})$, whereas a smaller $\sigma$ allow one to stay closer to the original spectral density $\phi$. To quantify the latter, one can measure the error between $\phi$ and its smoothened version $\phi_{\sigma}$ by
\begin{equation}
    \sup_{f \in \mathcal{S}} \int_{a}^{b} f(t) (\phi(t) - \phi_{\sigma}(t))~\mathrm{d}t,
    \label{equ:error-metric}
\end{equation}
where $\mathcal{S}$ represents an appropriately chosen space of test functions. Among others, the reference~\cite{lin-2016-approximating-spectral} uses $\mathcal{S} = \{ f: f(t) \equiv g_{\sigma}(t - s), s \in [a, b]\}$, \cite{chen-2021-analysis-stochastic} uses $\mathcal{S} = \{f : f(t) = \Theta(s - t), s \in [a, b] \}$ with the Heaviside step function $\Theta$, and \cite{braverman-2022-sublinear-time,bhattacharjee-2025-improved-spectral} use $\mathcal{S} = \{f : |f(t) - f(s)| \leq |t - s| \}$ for which the error \refequ{equ:error-metric} is equivalent to the Wasserstein-1 distance between $\phi$ and $\phi_{\sigma}$.
When choosing 
the last metric,
assuming that the eigenvalues are somewhat uniformly distributed and $\sigma$ is rather small, a simple calculation allows one to conclude that the smoothing error \refequ{equ:error-metric} is of the order $\sigma$. Consequently, to ensure that the approximation deviates by at most a factor of $\varepsilon > 0$, one needs to choose $\sigma \approx \varepsilon$.

@@

\begin{lemma}{Chebyshev approximation error}{chebyshev-error}
    The Chebyshev approximation $g_{\sigma}^{(m)}$ of the Gaussian smoothing kernel $g_{{\sigma}}$ from  \refequ{equ:smoothing-kernel} satisfies
    \begin{equation}
        \sup_{s,t \in [-1, 1]} \big| g_{\sigma}(t - s) - g_{\sigma}^{(m)}(t - s) \big| \leq 2\sqrt{\frac{2e}{\pi}} \frac{1}{\sigma^2} (1 + \sigma)^{-m} =: E_{\sigma, m}
        \label{equ:chebyshev-interpolation-sup-error-kernel}
        %\lVert  \phi_{\sigma} - \phi_{\sigma}^{(m)} \rVert _1 &\leq \frac{2\sqrt{2}}{\sigma^2} (1 + \sigma)^{-m}.
        %\label{equ:2-chebyshev-interpolation-error}
    \end{equation}
    for every degree $m \in \mathbb{N}$ and smoothing parameter $\sigma > 0$.
\end{lemma}
\begin{proof}
    The proof of this result is standard. % and follows the one given in the arXiv preprint of \cite[Theorem 2]{lin-2017-randomized-estimation}, but attains a slightly different dependence on $\sigma$. 
    Let $\mathcal{E}_{\chi} \subset \mathbb{C}$ denote the Bernstein ellipse  with foci $\{-1, +1\}$ and the sum of semi-axes equal to $\chi > 1$. By\cite[Theorem 8.2]{trefethen-2020-approximation-theory},
    \begin{equation*}
        \sup_{s \in [-1, 1]} \big| g_{\sigma}(t - s) - g_{\sigma}^{(m)}(t - s) \big| \leq \frac{4}{\chi^m (\chi - 1)} \max_{z \in \mathcal{E}_{\chi}} |g_{\sigma}(t - z)|,
        \label{equ:bernstein-bound}
    \end{equation*}
    where used the analyticity of $g_{\sigma}(t- \cdot)$.
    For $z = x + \mathrm{i} y \in \mathcal{E}_{\chi}$, we have
    \begin{equation*}
    | g_{\sigma}(t - z) | 
    = \frac{1}{\sigma \sqrt{2 \pi}} \big| e^{-\sfrac{(t - z)^2}{2\sigma^2}} \big|
    = \frac{1}{\sigma \sqrt{2 \pi}} e^{-\sfrac{(t - x)^2}{2\sigma^2}}e^{\sfrac{y^2}{2\sigma^2}}
    \le \frac{1}{\sigma \sqrt{2 \pi}} e^{\sfrac{(\chi - \chi^{-1})^2}{8 \sigma^2}},
    %\leq \frac{1}{n \sigma \sqrt{2 \pi}} \max_{x + iy \in \mathcal{E}_{\chi}} e^{\sfrac{y^2}{2\sigma^2}} 
    \end{equation*}
    where the last inequality uses 
    $e^{-\sfrac{(t - x)^2}{2\sigma^2}} \leq 1$ and $|y| \le (\chi - \chi^{-1}) / 2$.
    Choosing $\chi = 1 + \sigma$ and using $\chi - \chi^{-1} \leq 2\sigma$ completes the proof of~\refequ{equ:chebyshev-interpolation-sup-error-kernel}.
\end{proof}

@@

\begin{lemma}{$L^1$-error of non-negative approximate smooth spectral density}{non-negative-chebyshev-error}
Let $\phi_{\sigma}$ be the smoothed spectral density defined in~\refequ{equ:smooth-spectral-density}
for a symmetric matrix $\mtx{A} \in \mathbb{R}^{n \times n}$ with spectrum contained in $[-1, 1]$. Then the approximation $\underline{\phi}_{\sigma}^{(m)}$ defined in~\refequ{equ:non-negative-expanded-density} satisfies
    \begin{equation*}
        \int_{-1}^{1} \big| \phi_{\sigma}(t) - \underline{\phi}_{\sigma}^{(m)}(t) \big|~\mathrm{d}t \leq 4 \sqrt{2} \left(1 + \sigma \sqrt{\pi} \cdot E_{\sqrt{2}\sigma, \sfrac{m}{2}}\right) E_{\sqrt{2}\sigma, \sfrac{m}{2}} =: \underline{E}_{\sigma, m}
        \label{equ:chebyshev-interpolation-sup-error-kernel-nonneg}
        %\lVert  \phi_{\sigma} - \phi_{\sigma}^{(m)} \rVert _1 &\leq \frac{2\sqrt{2}}{\sigma^2} (1 + \sigma)^{-m}.
        %\label{equ:2-chebyshev-interpolation-error}
    \end{equation*}
    for every \emph{even} degree $m \in \mathbb{N}$, smoothing parameter $\sigma > 0$, and with $E_{\sigma, m}$ defined in \reflem{lem:chebyshev-error}.
\end{lemma}
\begin{proof}
    Using that the inequality $| a - b | \leq (2 \sqrt{a} + | \sqrt{a} - \sqrt{b} |)  | \sqrt{a} - \sqrt{b} |$ holds for any non-negative numbers $a, b$, it follows that
    \begin{equation*}
        \big| g_{\sigma} - \underline{g}_{\sigma}^{(m)} \big| \leq \big( 2 \big| \sqrt{g_{\sigma}} \big| + \big| \sqrt{g_{\sigma}} - \sqrt{g_{\sigma}}^{(\sfrac{m}{2})} \big| \big) \big| \sqrt{g_{\sigma}} - \sqrt{g_{\sigma}}^{(\sfrac{m}{2})} \big|.
    \end{equation*}
    Because of $\sqrt{g_{\sigma}} = \sqrt{2 \sigma \sqrt{2 \pi}} \cdot g_{\sqrt{2}\sigma}$,
we can apply \reflem{lem:chebyshev-error} with the substitutions
    $\sigma \gets \sqrt{2}\sigma$ and $m \gets m/2$ to bound 
    $\big| \sqrt{g_{\sigma}} - \sqrt{g_{\sigma}}^{(\sfrac{m}{2})} \big|$.
    Together with  $|\sqrt{g_{\sigma}}| \leq 1/\sqrt{\sigma \sqrt{2 \pi}}$, we get
    \begin{align*}
        \big| g_{\sigma}(t - s) - \underline{g}_{\sigma}^{(m)}(t - s) \big| \notag 
        &\leq \Big( 2 \frac{1}{\sqrt{ \sigma \sqrt{2\pi}}} + \sqrt{2 \sigma \sqrt{2 \pi}} \cdot E_{\sqrt{2}\sigma, \sfrac{m}{2}}\Big)\sqrt{2 \sigma \sqrt{2 \pi}} \cdot E_{\sqrt{2}\sigma, \sfrac{m}{2}} \notag \\
        &= 2\sqrt{2} \big(1 + \sigma \sqrt{\pi} \cdot E_{\sqrt{2}\sigma, \sfrac{m}{2}}\big) E_{\sqrt{2}\sigma, \sfrac{m}{2}}
    \end{align*}
    for every $s,t \in [-1, 1]$.
    Bounding the integral as in~\refequ{equ:chebyshev-interpolation-spectral-density} then yields the desired result.
\end{proof}

@@

%\textcolor{red}{??? I don't understand this derivation. Don't you think that~\refequ{equ:discrete-cosine-transform} only holds approximately for the exact $g_\sigma$ and it should be something like $g_\sigma \approx g^{(m)} = ...$. I think there is a major confusion of two different approaches here:
%\begin{itemize}
% \item The Chebyshev \emph{approximation} of $f$ is the $\ell^2$-orthogonal projection of $f$ on the first $m+1$ Chebyshev polynomials. The coefficients are computed from $\ell^2$ inner products between the Chebyshev polynomials and $f$. This is probably what Lin is doing. 
% \item The Chebyshev \emph{approximation} used by Chebfun and possibly others proceeds differently. It first replaces (or rather pretends to replace) $f$ by the interpolation of $f$ in (some) Chebyshev nodes. This interpolation admits an exact truncated approximation and the coefficients can be computed from a Fourier transform (or even DCT?) of the interpolation data. This might match what is written below, but I am not sure whether the interpolation nodes are the same. Please check with Trefethen's book.
%\end{itemize}
%There are two issues now
%\begin{enumerate}
% \item I am now unsure how $g_\sigma^{(m)}$ is precisely defined. This should be spelled out clearer in the beginning of Section 3.2.
% \item There could be a mismatch between the $g_\sigma^{(m)}$ of Lemma 3.1 and the one we are actually using. This mismatch can most likely be resolved, because a result of the form of Lemma 3.1 also holds for the Chebyshev interpolation. If it is easy to resolve, please do so. If not, at least point this out here. Have a look at Trefethen's book; it might have the right error bound.
%\end{enumerate}
%???
It is well known that the coefficients of the Chebyshev approximation \refequ{equ:matrix-approximation} can --- in exact arithmetic --- be computed by inverting the (type I) discrete cosine transform (DCT)
\begin{equation*}
    g_{\sigma}(t - \cos(s_i)) = \sum_{l=0}^{m} \mu_l(t) \cos\left(\frac{\pi i l}{m} \right)
    \label{equ:discrete-cosine-transform}
\end{equation*}
between the approximation coefficients $\mu_l(t),~l=0,\dots,m$ and the function evaluations at the Chebyshev points $s_i = \cos(\pi i / m),~i=0,\dots,m$ \cite{baszenski-1997-fast-polynomial, plonka-2018-numerical-fourier, trefethen-2020-approximation-theory}. This can be accomplished in $\mathcal{O}(m \log(m))$ operations.%}

@@

We start with the first term in the definition~\refequ{equ:nystrompp-trace-estimator} of $\Nystrpp{\mtx{\Omega}}{\mtx{\Psi}}$, which requires evaluating the trace of the 
Nyström approximation \refequ{equ:nystrom-approximation}:
\begin{equation}
    \frac{1}{n} \Trace \big((\mtx{B}(t) \mtx{\Omega}) (\mtx{\Omega}^{\top} \mtx{B}(t) \mtx{\Omega})^{\dagger} (\mtx{B}(t) \mtx{\Omega})^{\top} \big) =
    \frac{1}{n} \Trace \big((\underbrace{\mtx{\Omega}^{\top} \mtx{B}(t) \mtx{\Omega}}_{=: \mtx{K}_1(t)})^{\dagger} (\underbrace{\mtx{\Omega}^{\top} \mtx{B}(t)^2 \mtx{\Omega}}_{=: \mtx{K}_2(t)}) \big).
    %\frac{1}{n} \Trace((\mtx{\Omega}^{\top} \underline{g}_{\sigma}^{(m)}(t \mtx{I}_n - \mtx{A}) \mtx{\Omega})^{\dagger} (\mtx{\Omega}^{\top} \underline{g}_{\sigma}^{(m)}(t \mtx{I}_n - \mtx{A})^2 \mtx{\Omega})),
    \label{equ:cyclic-property}
\end{equation}
where we used the invariance of the trace under cyclic permutations.
Inserting the Chebyshev approximation  $\mtx{B}(t) = \underline{g}_{\sigma}^{(m)}(t \mtx{I}_n - \mtx{A})$ gives
\begin{equation*}
    \mtx{K}_1(t) = \mtx{\Omega}^{\top} \underline{g}_{\sigma}^{(m)}(t \mtx{I}_n - \mtx{A}) \mtx{\Omega} = \sum_{l=0}^{m} \underline{\mu}_l(t) (\mtx{\Omega}^{\top} T_l(\mtx{A}) \mtx{\Omega}).
\end{equation*}
Observing that $\mtx{\Omega}^{\top} T_l(\mtx{A}) \mtx{\Omega}$ is independent of $t$ allows us to precompute these (small) $n_{\mtx{\Omega}} \times n_{\mtx{\Omega}}$ matrices for $l=0, \dots, m$ using the Chebyshev recurrence \refequ{equ:chebyshev-recurrence}. For each value of $t$, we then only need to sum these matrices (multiplied by the corresponding coefficient $\underline{\mu}_l(t)$).

The squaring of Chebyshev approximations discussed in \refsec{sec:preservnonneg} allows us to express $\mtx{B}(t)^2 = \underline{g}_{\sigma}^{(m)}(t \mtx{I}_n - \mtx{A})^2 = \sum_{l=0}^{2m} \underline{\nu}_l(t) T_l(\mtx{A})$, where the coefficients $\underline{\nu}_l(t)$ can be cheaply computed from $\underline{\mu}_l(t)$ using~\refalg{alg:chebyshev-squaring}. Hence, if we precompute
$\mtx{\Omega}^{\top} T_l(\mtx{A}) \mtx{\Omega}$
also for $l=m+1, \dots, 2m$, this allows us to evaluate $\mtx{K}_2(t)$ cheaply, in an analogous way to $\mtx{K}_1(t)$. Alternatively, \cite{lin-2017-randomized-estimation} proposes to instead approximate the squared smoothing kernel $g_{\sigma}^2$ in a separate "auxiliary" Chebyshev series truncated after term $2m$. While this direct approximation of $g_{\sigma}^2$ might be more accurate than our squaring approach, it is \emph{inconsistent} in the sense that numerically computing the truncated Chebyshev series of $g_{\sigma}^2$ gives --- in general --- not the same result as squaring the truncated Chebyshev series of $g_{\sigma}$. We have observed that meticulously ensuring consistency between the approximations of $g_{\sigma}$ and $g_{\sigma}^2$ makes a visible difference in terms of accuracy (cf. \reffig{fig:interpolation-issue}). This consistency will also allow us to derive bounds for the estimator \refequ{equ:chebyshev-nystrom-formula}.
\begin{figure}[ht]
    \centering
    \input{plots/interpolation.pgf}
    \caption{Accuracy when computing the estimate \refequ{equ:cyclic-property} in three different ways. \emph{Inconsistent}: Applying an auxiliary truncated Chebyshev series to $g_{\sigma}^2$, as proposed in~\cite{lin-2017-randomized-estimation}. \emph{Consistent}: Squaring the Chebyshev approximation $g_{\sigma}^{(m)}$ using \refalg{alg:chebyshev-squaring}. \emph{Non-negative consistent}: Squaring the non-negative Chebyshev approximation $\underline{g}_{\sigma}^{(m)}$ with~\refalg{alg:chebyshev-squaring}. As input matrix we use the $100 \times 100$ matrix $\mtx{A}$ coming from a two-dimensional finite difference discretization of the Hamiltonian defined in~\refequ{equ:electronic-hamiltonian} on a $10 \times 10$ grid. We fix $n_{\mtx{\Omega}} = 80$ to make sure the low-rank approximation is accurate, set $\sigma = 0.005$, and compute the $L^1$ error from the approximation of the spectral density at $n_t = 100$ uniformly spaced values of $t$ in the interval $[-1, 1]$.}
    \label{fig:interpolation-issue}
\end{figure}

@@

\begin{figure}[ht]
    \centering
    \input{plots/zerocheck.pgf}
    \caption{The difference the non-zero check can make when approximating a spectral density using \refalg{alg:nystrom-chebyshev-pp}. As input matrix we use the $1000 \times 1000$ matrix $\mtx{A}$ coming from a three-dimensional finite difference discretization of the Hamiltonian defined in~\refequ{equ:electronic-hamiltonian} on a $10 \times 10$ grid. We ran the Chebyshev-Nyström++ method with and without estimating if the matrix function is zero before taking its pseudo-inverse. We let $m=2000$, $n_{\mtx{\Omega}}=80$, $n_{\mtx{\Psi}}=0$, and $\sigma = 0.004$. We compute the $L^1$ error from the approximation of the spectral density at $n_t = 100$ uniformly spaced values of $t$ in the interval $[-1, 1]$.}
    \label{fig:zerocheck}
\end{figure}
